---
title: "Manhattan-Taxis"
author: "Shreya Rao"
output:
  pdf_document: default
  html_document: default
---

Analyze analyze a random sample of 49999 New York journeys made in 2013 to help taxi drivers maximize thoer profits. Use regression trees and random forests to build a model that can predict the locations and times when the biggest fares can be earned.

```{r}
# Loading the tidyverse
library(tidyverse)

# Reading in the taxi data
taxi <- read.table("C:/Users/Shreya/Documents/Projects/taxi.txt", sep = ",", header = T)

# Taking a look at the first few rows in taxi
head(taxi)
```


Data Cleaning
```{r}
# Renaming the location variables,
# dropping any journeys with zero fares and zero tips,
# and creating the total variable as the log sum of fare and tip
taxi <- taxi %>%
   rename(lat = pickup_latitude, long = pickup_longitude) %>% 
filter(fare_amount > 0 | tip_amount > 0) %>% mutate(total = log(fare_amount + tip_amount))
```

```{r}
# Reducing the data to taxi trips starting in Manhattan
# Manhattan is bounded by the rectangle with 
# latitude from 40.70 to 40.83 and 
# longitude from -74.025 to -73.93
taxi <- taxi  %>% 
    filter(lat >= 40.70, lat <= 40.83, long >=  -74.025, long <= -73.93)
```

```{r}
# Loading in ggmap and viridis for nice colors
library(ggmap)
library(viridis)

register_google(key = "AIzaSyD1M0avrjJIOodaqQrRSs2qjBX6IlMksfI", write = TRUE)

# Retrieving a stored map object which originally was created by
manhattan <- get_map("manhattan", zoom = 12, color = "bw")
# <- readRDS("C:/Users/Shreya/Documents/Projects/manhattan.rds")

# Drawing a density map with the number of journey start locations
ggmap(manhattan, darken = 0.5) +
   scale_fill_viridis(option = 'plasma') +
   geom_bin2d(data = taxi, mapping = aes(x = long, y = lat), alpha = 0.6) + 
   labs(x = "Longitude", y = "Latitude")
```


**Predicting Taxi Fares Using a Tree**
```{r}
# Loading in the tree package
library(tree)

# Fitting a tree to lat and long
fitted_tree <- tree(total~lat+long, data=taxi)

# Draw a diagram of the tree structure
plot(fitted_tree)
text(fitted_tree)
```

It predicts that trips where lat < 40.7237 are more expensive, which makes sense as it is downtown Manhattan.

Adding some more predictors related to the time the taxi trip was made:
```{r}
# Loading in the lubridate package
library(lubridate)

# Generate the three new time variables
taxi <- taxi %>% 
    mutate(hour = hour(pickup_datetime), 
           wday = wday(pickup_datetime, label = TRUE), 
           month = month(pickup_datetime, label = TRUE))
```


```{r}
# Fitting a tree with total as the outcome and 
# lat, long, hour, wday, and month as predictors
fitted_tree <- tree(total~lat+long+hour+wday+month, data=taxi)

# draw a diagram of the tree structure
plot(fitted_tree)
text(fitted_tree)

# Summarizing the performance of the tree
summary(fitted_tree)
```


**Random Forest**
```{r}
# Loading in the randomForest package
library(randomForest)

# Fitting a random forest
fitted_forest <- randomForest(total~lat+long+hour+wday+month, data=taxi,
                             ntree = 80, sampsize = 10000)

# Printing the fitted_forest object
print(fitted_forest)
```


**Plotting the Predicted Fare**
```{r}
# Extracting the prediction from fitted_forest
taxi$pred_total <- fitted_forest$predicted

# Plotting the predicted mean trip prices from according to the random forest
ggmap(manhattan, darken = 0.5) +
   scale_fill_viridis(option = 'plasma') +
   stat_summary_2d(data = taxi, mapping = aes(x = long, y = lat, , z = pred_total), alpha = 0.6) + 
   labs(x = "Longitude", y = "Latitude", fill = "Predicted Fare")
```

<!-- **Plotting the Actual Fare** -->
<!-- ```{r} -->
<!-- # Function that returns the mean *if* there are 15 or more datapoints -->
<!-- mean_if_enough_data <- function(x) {  -->
<!--     ifelse(length(x) >= 15, mean(x), NA)  -->
<!-- } -->

<!-- # Plotting the mean trip prices from the data -->
<!-- ggmap(manhattan, darken = 0.5) + -->
<!--    scale_fill_viridis(option = 'plasma') + -->
<!--    stat_summary_2d(data = taxi, mapping = aes(x = long, y = lat, ,  -->
<!--                         z = total, fun = mean_if_enough_data), alpha = 0.6) +  -->
<!--    labs(x = "Longitude", y = "Latitude", fill = "Predicted Fare") -->
<!-- ``` -->

